---
title: "MovieLens Report"
author: "Zachary Naumann"
date: "4/15/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(recommenderlab)) install.packages("recommenderlab", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)
library(recommenderlab)

edx <- read_csv("edx.csv")
```

## Introduction
The purpose of this report was to determine the effect of various factors on
movie ratings. The relevant dataset was contained in a data frame with 6
variables, with each element representing an instance of a movie watched by a
user. Below is a sample of the dataset for reference:
```{r structure, echo=FALSE}
head(edx)
```
The purpose of each variable is as follows:  
1. *userId* - unique numeric identifier for the user  
2. *movieId* - unique numeric identifier for the movie  
3. *rating* - rating given by the user to the movie; the lowest rating
possible is 0.5, the highest possible is 5  
4. *timestamp* - timestamp for when the user watched the movie  
5. *title* - title of the movie  
6. *genres* - genres assigned to the movie; a movie can have 1 or more genres  
  
  
There were several key steps required to gather the necessary data.

## Analysis

## Results

## Conclusion